Gravitational wave sources in our Galactic backyard: Predictions for BH and NS binaries in LISA

### 2021-07-07 paper/tex/sections/results.tex
Lieke section 4

\begin{figure*}[p]
    \centering
    \includegraphics[width=\textwidth]{2_dcos_on_sc.png}
    \caption{Density distribution of detectable BHBH, BHNS and NSNS binaries are shown together with the LISA sensitivity curve. In the top panel we show all systems with the LISA verification binaries over plotted (star symbols, \citealp{Kupfer+2018}). In the bottom panels we separate by type. Contours show the percentage of the population enclosed. The remaining 2\% of the population is shown as dots with a size that scales with the sampling weight. For reference we show lines where a circular binary of average mass $\avg{\mathcal{M}_c}$ would reside for a given remaining inspiral time (vertical lines) and distance (diagonal line). To highlight the role of eccentricity we further show the signal expected for an eccentric binary at 30 kpc. The coloured line in the bottom panels shows a contour that encloses 90\% of the population that is circular. See Sec.~\ref{sec:dcos_on_sc} for a discussion.}
    \label{fig:dcos_on_sc}
\end{figure*}

@@

Firstly, we note that the peak of the density distribution coincides with the centre of the Milky Way as expected, since binaries are most likely to be formed towards the centre of the Galaxy. Additionally, we expect that if a population is entirely circular, it should be bounded approximately between the $0.1$--$30 \unit{kpc}$ lines, roughly the minimum and maximum distance to a source in the Milky Way. From inspection of the bottom panels with each individual DCO type we see that this is the true for a large fraction of the population. However, there is a distinct subpopulation of binaries that extend downwards, especially around $2 \unit{mHz}$. This offshoot is composed of eccentric binaries for which the circular distance contours do not apply. For instance, we plot the 90\% contour of only the \textit{circular} sources in our sample over the density distribution in each of the bottom panels and it is indeed bounded by $0.1$--$30 \unit{kpc}$ the distance lines. We also plot a line of constant distance at $30 \unit{kpc}$ for an eccentric binary with $e = 0.98$ to show the different limit for eccentric sources. To further illustrate this point, we show the same plot but with each point coloured by its eccentricity in Fig.~\ref{fig:dcos_on_sc_ecc_col}. Overall, we can see that eccentric sources tend to be located on a slightly different location on the sensitivity curve and so this could perhaps be used help in identifying them.

@@

\subsubsection{Black Hole Mass}
For both the BHBHs and BHNSs, the black hole mass distribution extends across relatively low masses, with $88\%$ and $90\%$ respectively below $11 \unit{M_{\odot}}$. Unlike ground-based detector, LISA is not biased to higher masses and so the mass distribution more closely follows the IMF. In addition, at the high metallicities in the Milky Way, stellar winds are much stronger and strip away much of the stellar mass before BH formation, resulting the less massive black holes. The mass distribution extends down to $2.5 \unit{M_{\odot}}$, our fiducial maximum neutron star mass, since the \citet{Fryer+2012} \textit{delayed} remnant mass prescription does not produce a mass gap between neutron stars and black holes. Indeed we expect $35\%$ and $39\%$ of detected BHBH and BHNS systems to contain a black hole in the lower mass gap. Therefore, LISA could help to confirm or rule out the existence of the lower mass gap.

The bimodality of the BHBH distribution is a result of most detectable BHBHs in our sample having unequal mass ratios. The two peaks are from the primary and secondary black hole masses, which peak around $8 \unit{M_{\odot}}$ and $3.5 \unit{M_{\odot}}$ respectively. We show these individual distributions as dotted curves below the main BHBH distribution.

@@

\subsubsection{Neutron Star Mass}
The neutron star mass distribution shows that most neutron stars have low masses, with $77\%$ and $91\%$ having masses below $1.7 \unit{M_{\odot}}$ for BHNSs and NSNSs respectively. The lack of neutron stars around $1.7 \unit{M_{\odot}}$ and the subsequent small peaks are artifacts of the discontinuous nature of the \citet{Fryer+2012} remnant mass prescription and do not have strong effects since the vast majority of NSs are formed with lower masses.

Both distributions have notable peaks around $1.26 \unit{M_{\odot}}$, though more strongly in the NSNS case, which are a result of the combination of two effects. Firstly, we set the remnant mass for all electron-capture supernovae to $1.26 \unit{M_\odot}$ (see Sec.~\ref{sec:fiducial_physics}) and thus this leads to a pileup when many systems are formed through ECSN. In addition, the Fryer remnant mass prescription gives a fixed fallback mass for any star with a CO core mass less than $2.5 \unit{M_\odot}$, such that many NSs are given the identical mass of $1.278 \unit{M_\odot}$ in the \textit{delayed} prescription \citep[see][Eq.~19]{Fryer+2012}.

\subsubsection{Eccentricity}
The eccentricity distributions show that detectable BHBHs are the most likely to be highly eccentric of the three DCOs. This may seem counter-intuitive since neutron stars receive stronger natal kicks, which cause the orbit to become eccentric. However, these stronger kicks often instead result in disrupted or too-wide binaries in more weakly bound NSNSs. In contrast, BHBHs can receive strong kicks that impart high eccentricity without disrupting and thus tend to be more eccentric. This effect is compounded by the fact that we can see BHBHs at lower orbital frequencies, meaning that they have not had as much time to circularise and so still have significant eccentricity by the time of the LISA mission.

An eccentricity of $e = 0.01$ is the lower bound on the measurable eccentricity with LISA proposed by \citet{Nishizawa+2016}. We find that a significant fraction of DCOs (86\% of BHBHs, 46\% of BHNSs and 86\% of NSNSs) exceed this bound. This is in contrast to several previous work that assume all DCOs are circular once they reach the LISA mission. Moreover, we find that 20\%, 8\% and 12\% of BHBHs, BHNSs and NSNSs respectively have $e > 0.3$. At this eccentricity the majority of the gravitational wave power is emitted in higher harmonics above the orbital frequency and thus the signal would appear at higher frequencies in LISA.

@@

However, more surprisingly, we also see that the NSNS distribution extends to higher distances than the other DCOs. The reason for this is that the NSNS population has the highest fraction of ``mildly'' eccentric systems ($0.01 < e < 0.3$). The BHNS population has a much higher fraction of effectively circular systems ($e < 0.01$), which emit weaker gravitational wave compared to equivalent eccentric systems. Therefore, despite their relatively higher masses, the maximum distance at which a source is detectable is generally lower than the eccentric NSNSs and hence the distribution tends to zero at smaller distances. Conversely, the BHBH population has a higher fraction of \textit{high} eccentricity systems $e > 0.3$. Although one may naively expect that this would result in stronger signals (and so further distances), for a system to have these high eccentricities in LISA it must still be early in its evolution and thus have a low orbital frequencies. The result of this is that high eccentricity systems tend to have lower SNRs and so cannot be detected at large distances. Overall we see that the eccentricity distribution of NSNSs occupies a ``sweet spot'' where the gravitational wave power is increased compared to circular systems, but it isn't too high that the frequency is significantly impacted. This means that NSNSs can be seen at the largest distances of the three DCO types.

@@

The chirp mass uncertainty is important for identifying the source of a GW signal. It is calculated using the uncertainty on the orbital frequency, the time derivative of the orbital frequency and the eccentricity. This is because the chirp mass can be written as
\begin{equation}
    \mathcal{M}_c = \frac{c^3}{G} \left( \frac{5 \pi}{48 n} \frac{\dot{f}_{n}}{F(e)} \right)^{3/5} \frac{1}{(2 \pi f_{\rm orb})^{11/5}},
\end{equation}
where $f_{\rm orb}$ is the orbital frequency, $\mathcal{M}_{c}$ is the chirp mass (defined in Eq.~\ref{eq:chirp_mass}) and $e$ is the eccentricity and
\begin{equation}
    F(e) = \frac{1 + \frac{73}{24} e^2 + \frac{37}{96} e^4}{(1 - e^2)^{7/2}},
\end{equation}
is the enhancement factor of gravitational wave emission for an eccentric binary over an otherwise identical circular binary \citep[][Eq.~17]{Peters+1963}. Therefore that the chirp mass uncertainty is
\begin{equation}\label{eq:chirp_mass_uncertainty}
    \qty( \frac{\Delta \mathcal{M}_c}{\mathcal{M}_c} )^2 = \left( \frac{11}{5} \frac{\Delta f_{\rm orb}}{f_{\rm orb}} \right)^2 + \left( \frac{3}{5} \frac{\Delta \dot{f}_{\rm dom}}{\dot{f}_{\rm dom}} \right)^2 + \left( \frac{3}{5} \frac{\Delta F(e)}{F(e)} \right)^2,
\end{equation}
where $f_{\rm dom}$ is the harmonic frequency with the strongest SNR ($f_{\rm dom} = n_{\rm dom} f_{\rm orb}$ and $n_{\rm dom} = 2$ for circular binaries) as this will provide the best measurement.

@@

We use Eq.~\ref{eq:chirp_mass_uncertainty} to calculate the chirp mass uncertainty for each DCO in our sample and plot it in Fig.~\ref{fig:m_c_unc}. We find that approximately 10 BHBHs, 10 BHNSs and 5 NSNSs have measurable chirp masses (as indicated by the shaded region). This uncertainty is generally dominated by the uncertainty on the time derivative of the frequency since most of the binaries are too early in their inspiral for LISA to measure a strong chirp.

>>>>

\begin{figure*}[p]
    \centering
    \includegraphics[width=\textwidth]{2_dcos_on_sc.png}
    \caption{Density distribution of detectable BHBH, BHNS and NSNS binaries are shown together with the LISA sensitivity curve. In the top panel we show all systems with the LISA verification binaries over plotted (star symbols, \citealp{Kupfer+2018}). In the bottom panels we separate by type. Contours show the percentage of the population enclosed. The remaining 2\% of the population is shown as dots with a size that scales with the sampling weight. For reference we show lines where a circular binary of average chirp mass $\avg{\mathcal{M}_c}$ would reside for a given remaining inspiral time (vertical lines) and distance (diagonal line). To highlight the role of eccentricity we further show the signal expected for an eccentric binary at 30 kpc. The coloured line in the bottom panels shows a contour that encloses 90\% of the population that is circular. See Sec.~\ref{sec:dcos_on_sc} for a discussion.}
    \label{fig:dcos_on_sc}
\end{figure*}

@@

Firstly, we note that the peak of the density distribution coincides with the centre of the Milky Way as expected, since binaries are most likely to be formed towards the centre of the Galaxy. Additionally, we expect that if a population is entirely circular, it should be bounded approximately between the $0.1$--$30 \unit{kpc}$ lines, roughly the minimum and maximum distance to a source in the Milky Way. From inspection of the bottom panels with each individual DCO type we see that this is the true for a large fraction of the population. However, there is a distinct subpopulation of binaries that extend downwards, especially around $2 \unit{mHz}$. This offshoot is composed of eccentric binaries for which the circular distance contours do not apply. For illustrative purposes, we plot the 90\% contour of only the \textit{circular} sources in our sample over the density distribution in each of the bottom panels and this subsample is indeed bounded by $0.1$--$30 \unit{kpc}$ the distance lines. We also plot a line of constant distance at $30 \unit{kpc}$ for an eccentric binary with $e = 0.98$ to show the different limit for eccentric sources. To further illustrate this point, we show the same plot but with each point coloured by its eccentricity in Fig.~\ref{fig:dcos_on_sc_ecc_col}. Overall, we can see that eccentric sources tend to be located on a slightly different location on the sensitivity curve and so this could perhaps be used help in identifying them.

@@

\subsubsection{Black Hole Mass}
For both the BHBHs and BHNSs, the black hole mass distribution extends to relatively low masses, with $88\%$ and $90\%$ respectively below $11 \unit{M_{\odot}}$. Unlike ground-based detectors, LISA is not biased to higher masses and so the mass distribution more closely follows the IMF. In addition, at the high metallicities in the Milky Way, stellar winds are much stronger and strip away much of the stellar mass before BH formation, resulting the less massive black holes. The mass distribution extends down to $2.5 \unit{M_{\odot}}$, our fiducial maximum neutron star mass. Since the \citet{Fryer+2012} \textit{delayed} remnant mass prescription (applied in this work) does not produce a mass gap between neutron stars and black holes. Indeed we expect $35\%$ and $39\%$ of detected BHBH and BHNS systems to contain a black hole in the lower mass gap. Therefore, LISA could help to confirm or rule out the existence of the lower mass gap.

The bimodality of the BHBH distribution is a result of unequal mass ratios. The two peaks are from the primary and secondary black hole masses, which peak around $8 \unit{M_{\odot}}$ and $3.5 \unit{M_{\odot}}$ respectively. We show these individual distributions as dotted curves below the main BHBH distribution.

@@

\subsubsection{Neutron Star Mass}
The neutron star mass distribution shows that most neutron stars have low masses, with $77\%$ and $91\%$ having masses below $1.7 \unit{M_{\odot}}$ for BHNSs and NSNSs respectively. The lack of neutron stars around $1.7 \unit{M_{\odot}}$ and the subsequent small peaks are a result of the discontinuous nature of the \citet{Fryer+2012} remnant mass prescription but do not have strong effects on our results since the vast majority of NSs are formed with lower masses.

Both distributions have a notable peak around $1.26 \unit{M_{\odot}}$, which is a result of the combination of two effects. Firstly, we set the remnant mass for all electron-capture supernovae to $1.26 \unit{M_\odot}$ (see Sec.~\ref{sec:fiducial_physics}) and thus this leads to a pileup when many systems are formed through ECSN. In addition, the Fryer remnant mass prescription gives a fixed fallback mass for any star with a CO core mass less than $2.5 \unit{M_\odot}$, such that many NSs are given the identical mass of $1.278 \unit{M_\odot}$ in the \textit{delayed} prescription \citep[see][Eq.~19]{Fryer+2012}.

\subsubsection{Eccentricity}
The eccentricity distributions show that detectable BHBHs are most likely to be highly eccentric of the three DCOs. This may seem counter-intuitive since neutron stars receive stronger natal kicks, which cause the orbit to become eccentric. However, these stronger kicks often result in disrupted or too-wide binaries in the more weakly bound NSNSs. In contrast, BHBHs can receive strong kicks that impart high eccentricity without disrupting and thus tend to be more eccentric. This effect is compounded by the fact that we can see BHBHs at lower orbital frequencies, meaning that they have not had as much time to circularise and so still have significant eccentricity by the time of the LISA mission.

An eccentricity of $e = 0.01$ is the lower bound on the measurable eccentricity with LISA proposed by \citet{Nishizawa+2016}. We find that a significant fraction of DCOs (86\% of BHBHs, 46\% of BHNSs and 86\% of NSNSs) exceed this bound. This is in contrast to several previous work that assume all DCOs are circular once they reach the LISA mission \citep[e.g.][]{Lamberts+2018, Sesana+2020}. Moreover, we find that 20\%, 8\% and 12\% of BHBHs, BHNSs and NSNSs respectively have $e > 0.3$. At this eccentricity the majority of the gravitational wave power is emitted in higher harmonics above the orbital frequency and thus the signal would appear at higher frequencies in LISA.

@@

However, more surprisingly, we also see that the NSNS distribution extends to higher distances than the other DCOs. The reason for this is that the NSNS population has the highest fraction of ``mildly'' eccentric systems ($0.01 < e < 0.3$). The BHNS population has a much higher fraction of effectively circular systems ($e < 0.01$), which emit weaker gravitational wave compared to equivalent eccentric systems. Therefore, despite their relatively higher masses, the maximum distance at which a source is detectable is generally lower than the eccentric NSNSs and hence the distribution tends to zero at smaller distances. Conversely, the BHBH population has a higher fraction of \textit{high} eccentricity systems $e > 0.3$. Although one may naively expect that this would result in stronger signals (and so further distances), for a system to have these high eccentricities in LISA, it must still be early in its evolution and thus have a low orbital frequencies. The result of this is that high eccentricity systems tend to have lower SNRs and so cannot be detected at large distances. Overall we see that the eccentricity distribution of NSNSs occupies a ``sweet spot'' where the gravitational wave power is increased compared to circular systems, but it isn't too high that the frequency is significantly impacted. This means that NSNSs can be seen out to the largest distances of the three DCO types.

@@

The chirp mass uncertainty is important for identifying the source of a GW signal. It is calculated using the uncertainty on the orbital frequency, the time derivative of the orbital frequency and the eccentricity. This is because the chirp mass can be written as
\begin{equation}
    \mathcal{M}_c = \frac{c^3}{G} \left( \frac{5 \pi}{48 n} \frac{\dot{f}_{n}}{F(e)} \right)^{3/5} \frac{1}{(2 \pi f_{\rm orb})^{11/5}},
\end{equation}
where $f_{\rm orb}$ is the orbital frequency, $\mathcal{M}_{c}$ is the chirp mass (defined in Eq.~\ref{eq:chirp_mass}) and $e$ is the eccentricity and
\begin{equation}
    F(e) = \frac{1 + \frac{73}{24} e^2 + \frac{37}{96} e^4}{(1 - e^2)^{7/2}},
\end{equation}
is the enhancement factor of gravitational wave emission for an eccentric binary over an otherwise identical circular binary \citep[][Eq.~17]{Peters+1963}. Therefore the chirp mass uncertainty is
\begin{equation}\label{eq:chirp_mass_uncertainty}
    \qty( \frac{\Delta \mathcal{M}_c}{\mathcal{M}_c} )^2 = \left( \frac{11}{5} \frac{\Delta f_{\rm orb}}{f_{\rm orb}} \right)^2 + \left( \frac{3}{5} \frac{\Delta \dot{f}_{\rm dom}}{\dot{f}_{\rm dom}} \right)^2 + \left( \frac{3}{5} \frac{\Delta F(e)}{F(e)} \right)^2,
\end{equation}
where $f_{\rm dom}$ is the harmonic frequency with the strongest SNR ($f_{\rm dom} = n_{\rm dom} f_{\rm orb}$ and $n_{\rm dom} = 2$ for circular binaries) as this will provide the best measurement.

@@

We use Eq.~\ref{eq:chirp_mass_uncertainty} to calculate the chirp mass uncertainty for each DCO in our sample and plot it in Fig.~\ref{fig:m_c_unc}. We find that approximately 10 BHBHs, 10 BHNSs and 5 NSNSs have measurable chirp masses (as indicated by the shaded region). This uncertainty is generally dominated by the uncertainty on the time derivative of the frequency, since most of the binaries are too early in their inspiral for LISA to measure a strong chirp.

### 2021-07-07 paper/tex/sections/variations.tex
Lieke section 4

In this section we explore the effect of varying our underlying physics assumptions and model for the Milky Way. In Section~\ref{sec:detection_rate_analysis} we explore the changes in the detection rate over different physics variations  and in Section~\ref{sec:property_variations} we consider how these model physics variations change the properties of detectable systems. Finally we demonstrate how different models for the Milky Way can affect our results in Section~\ref{sec:mw_changes}.

@@

Their model can be summarised as follows: the Milky Way is assumed to comprise of three components, a thin disc, a thick disc and a bulge. The spatial distributions and relative masses for these components are given in \citet{McMillan+2011}. \citet{Breivik+2020} assume constant star formation over 10 Gyr for the thin disc, a 1 Gyr burst of star formation 11 Gyr ago for the thick disc and a 1 Gyr burst of star formation 10 Gyr ago for the bulge. A major difference is that only two metallicities are used and they are assigned independent of age or position. Binaries formed in the thin disc and bulge are assumed to have a metallicity of $Z = 0.02$ and those formed in the thick disc are assumed to have $Z = 0.003$.

We show this model in Fig.~\ref{fig:simple_mw} in the same form as Fig.~\ref{fig:galaxy_schematic} for ease of comparison between our models. The two main differences we can see in these plots are that the \citet{Breivik+2020} model is more centrally concentrated and only has two fixed metallicity populations.
%In addition, though it can't be seen in these plots, the simpler prescriptions for the star formation histories mean that the overall star formation history is highly discontinuous at $\tau = 9 \unit{Gyr}$ and significantly more star formation occurs from $\tau = 9 \unit{Gyr}$ to $\tau = 11 \unit{Gyr}$.

@@

We find that when applying this simpler Milky Way model and our fiducial physics assumptions (model \modFid{}), the expected number of detections for BHBHs, BHNSs and NSNSs for a 4-year LISA mission is $39$, $39$ and $14$ respectively. Thus the BHBH and BHNS detection rates are only marginally increased from our main findings, but the NSNS detection is overestimated by nearly a factor of 2.

>>>>

In this section we explore the effect of varying the underlying binary physics assumptions and the model for the Milky Way. In Section~\ref{sec:detection_rate_analysis} we explore the changes in the detection rate over different physics variations  and in Section~\ref{sec:property_variations} we consider how these model physics variations change the properties of detectable systems. Finally we demonstrate how different models for the Milky Way can affect our results in Section~\ref{sec:mw_changes}.

@@

Their model can be summarised as follows: the Milky Way is assumed to comprise of three components, a thin disc, a thick disc and a bulge. The spatial distributions and relative masses for these components are given in \citet{McMillan+2011}. \citet{Breivik+2020} assume constant star formation over 10 Gyr for the thin disc, a 1 Gyr burst of star formation 11 Gyr ago for the thick disc and a 1 Gyr burst of star formation 10 Gyr ago for the bulge. A major difference is that only two metallicities are used and they are assigned to binaries independent of age or position. Binaries formed in the thin disc and bulge are assumed to have a metallicity of $Z = 0.02$ and those formed in the thick disc are assumed to have $Z = 0.003$.

We show the spatial metallicity distribution for this model in Fig.~\ref{fig:simple_mw} in the same form as Fig.~\ref{fig:galaxy_schematic} for ease of comparison between our models. The two main differences we can see between Fig.~\ref{fig:galaxy_schematic} and \ref{fig:simple_mw} are that the \citet{Breivik+2020} model is more centrally concentrated and only has two fixed metallicity populations.
%In addition, though it can't be seen in these plots, the simpler prescriptions for the star formation histories mean that the overall star formation history is highly discontinuous at $\tau = 9 \unit{Gyr}$ and significantly more star formation occurs from $\tau = 9 \unit{Gyr}$ to $\tau = 11 \unit{Gyr}$.

@@

When applying this simpler Milky Way model in combination with our fiducial binary physics assumptions (model \modFid{}), we find that the expected number of detections for BHBHs, BHNSs and NSNSs for a 4-year LISA mission is $39$, $39$ and $14$ respectively. Thus the BHBH and BHNS detection rates are only marginally increased from our main findings, but the NSNS detection has increased by nearly a factor of 2.
